---
title: "Energy Flows from the IM3 GCAM-USA Scenarios"
author: "Hassan Niazi (hassan.niazi@pnnl.gov) | Adapted from the work of Rachel Hoesly"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    toc: true
    # toc_float: TRUE
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE, warning=FALSE}
# by default collapse/hide the code
# knitr::opts_chunk$set(echo = FALSE)
# set working directory to one folder up
setwd("../")
# getwd()
source("./R/functions.R")
```

### For the IWPR Q1 Update (Jan 2025)

-   Goal: plot an all energy sankey for Q1 update of the EW-Flows project

### Query and load IM3 data

Run this `query_im3_scen("energy")` only once to query from remote IM3 databases. Once a `.dat` file is created, we can load the existing project data by `loadProject(proj = "im3scen_energy.dat")`.

```{r message=FALSE, warning=FALSE}
# query the data
# im3_energy <- query_im3_scen("energy")
```

```{r, warning = FALSE}
# load the data
im3_energy <- loadProject(proj = paste0("../", data_dir, "im3scen_energy.dat"))
```

```{r}
# scenarios and queries 
listScenarios(im3_energy)
listQueries(im3_energy)
```


```{r, warning = FALSE}
# mappings 
source_mapping_e <- read_csv(paste0("../", data_dir, "mappings/source_mapping_e.csv"))
target_mapping_e <- read_csv(paste0("../", data_dir, "mappings/target_mapping_e.csv"))
node_mapping_e <- read_csv(paste0("../", data_dir, "mappings/node_mapping_e.csv")) 
```


### Energy Sankey

```{r, warning = FALSE}
# get queries 
inputsByTechUSA <- getQuery(im3_energy, "USA inputs by tech") 
outputsByTechUSA <- getQuery(im3_energy, "USA outputs by tech")

inputBySubsectorNonElec <- getQuery(im3_energy, 'inputs by subsector (non-electric)') %>% filter_CONUSregions() 
elecEnergyInputBySubsector <- getQuery(im3_energy, 'elec energy input by subsector') %>% filter(Units == "EJ") %>% filter_CONUSregions() # in case no filtering of ELEC_RPS credits
elecGenBySubsector <- getQuery(im3_energy, 'elec gen by subsector') %>% filter(Units == "EJ") %>% filter_CONUSregions() # in case no filtering of ELEC_RPS credits
natGasOutputs <- getQuery(im3_energy, 'USA regional natural gas outputs')

```

```{r, messsage = T, warning = FALSE}
datatables_energy <- list(
  "inputsByTechUSA" = inputsByTechUSA,
  "outputsByTechUSA" = outputsByTechUSA,
  "inputBySubsectorNonElec" = inputBySubsectorNonElec,
  "elecEnergyInputBySubsector" = elecEnergyInputBySubsector,
  "elecGenBySubsector" = elecGenBySubsector,
  "natGasOutputs" = natGasOutputs
)

# print column names of each datatable
lapply(datatables_energy, function(x) colnames(x))

# print the first few rows of each datatable
lapply(datatables_energy, function(x) (x))
```

Let's process each piece to prepare the format of: scenario, source, target, year, value. Scenario and year could be filtered for each Sankey.

#### Non-electricity 


```{r fig.width=8, warning=FALSE}
unique((inputBySubsectorNonElec %>% remove_month_day_night_superpeak("sector"))$sector)
unique(inputBySubsectorNonElec$subsector)
unique((inputBySubsectorNonElec %>% remove_month_day_night_superpeak("input"))$input)
```


```{r fig.width=8, warning=FALSE}
# map non electricity energy flows to major aggregated categories based on the mapping file 

inputs_by_subsector_nonelec <- inputBySubsectorNonElec %>% 
  filter(Units == 'EJ') %>% # only take energy flows
  filter(!input %in% c('regional corn', 'regional soybean')) %>% # remove crop inputs 
  # aggregate all monthly_day combinations to one category e.g., electricity domestic supply_Nov_day to electricity domestic supply
  remove_month_day_night_superpeak("sector") %>% remove_month_day_night_superpeak("input") %>%
  rbind(inputsByTechUSA %>% filter(str_detect(sector, "H2 ")) %>% select(-technology)) %>% # add H3 flows from the USA region level as IM3 doesn't model H2 at state level
  left_join(source_mapping_e, by = 'input') %>%
  left_join(target_mapping_e, by = 'sector')

# Note there are NAs in the output due to missing mappings or sectors that are
# not supposed to be targets and inputs that are not supposed# to be sources
```


```{r fig.width=8, warning=FALSE}
# get hydrogen flows from USA region since IM3 version doesn't model H2 at state level 
inputsByTechUSA %>%# filter all inputs that have "H2 " in it
  filter(str_detect(sector, "H2")) %>%
  select(sector) %>% unique()


```


```{r fig.width=8, warning=FALSE}
# things that were remapped as sources
unique((inputs_by_subsector_nonelec %>% filter(!is.na(Source)))$sector)
unique((inputs_by_subsector_nonelec %>% filter(!is.na(Source)))$subsector)
unique((inputs_by_subsector_nonelec %>% filter(!is.na(Source)))$input) # look at this
```

```{r fig.width=8, warning=FALSE}
# things there were NOT mapped as sources 
unique((inputs_by_subsector_nonelec %>% filter(is.na(Source)))$sector)
unique((inputs_by_subsector_nonelec %>% filter(is.na(Source)))$subsector)
unique((inputs_by_subsector_nonelec %>% filter(is.na(Source)))$input) # look at this 
```


```{r fig.width=8, warning=FALSE}
# things that were remapped as targets
unique((inputs_by_subsector_nonelec %>% filter(!is.na(Target)))$sector) # look at this
unique((inputs_by_subsector_nonelec %>% filter(!is.na(Target)))$subsector)
unique((inputs_by_subsector_nonelec %>% filter(!is.na(Target)))$input)
```


```{r fig.width=8, warning=FALSE}
# things that were NOT remapped as targets
unique((inputs_by_subsector_nonelec %>% filter(is.na(Target)))$sector) # look at this
unique((inputs_by_subsector_nonelec %>% filter(is.na(Target)))$subsector)
unique((inputs_by_subsector_nonelec %>% filter(is.na(Target)))$input)
```


```{r fig.width=8, warning=FALSE}
# check for unmatched sources
inputs_by_subsector_nonelec_unmatched_source <- inputs_by_subsector_nonelec %>% 
  filter(is.na(Source)) %>% 
  select(scenario, sector, subsector, input, Source, Target) %>% 
  unique()

unique(inputs_by_subsector_nonelec_unmatched_source$input)

unmatched_sources <- c("oil-credits", "renewable", "regional oilcrop", "process heat cement", "process heat dac")

if(! all(inputs_by_subsector_nonelec_unmatched_source$input %in% unmatched_sources )){
  unmatched <- setdiff(inputs_by_subsector_nonelec_unmatched_source$input, unmatched_sources)
  stop(paste0("Unmatched Sources in inputs by subsector nonelec. Check Source mapping file against gcam data: ", paste(unmatched, collapse = ' - ')))
}
```


```{r fig.width=8, warning=FALSE}
# check for unmatched targets
inputs_by_subsector_nonelec_unmatched_target <- inputs_by_subsector_nonelec %>% 
  filter(is.na(Target)) %>% 
  select(scenario, sector, subsector, input, Source, Target) %>% 
  unique

unique(inputs_by_subsector_nonelec_unmatched_target$sector)

unmatched_targets <- c("H2 central production", 
                       "H2 liquid truck",
                       "H2 pipeline",
                       "H2 wholesale delivery" #all intermediate hydrogen markets that are double counting - only want H2 industrial and H2 MHDV
                       )
if(! all(inputs_by_subsector_nonelec_unmatched_target$sector %in% unmatched_targets)){
  unmatched <- setdiff(inputs_by_subsector_nonelec_unmatched_target$sector, unmatched_targets)
  stop(paste0("Unmatched Sources in inputs by subsector nonelec. Check Source mapping file against gcam data: ", paste(unmatched, collapse = ' - ')))
}


```

Get other flows such as gas processing and backup electricity

```{r fig.width=8, warning=FALSE}
gas_processing_flows <- inputsByTechUSA %>%
  filter(sector == "gas processing") %>%
  left_join(source_mapping_e, by = "input") %>%
  left_join(target_mapping_e, by = "sector") %>%
  group_by(scenario, Units, year, Source, Target) %>%
  summarize(value = sum(value)) %>%
  ungroup()

backup <- inputsByTechUSA %>%
  filter(sector %in% c("backup_electricity", "csp_backup")) %>%
  left_join(source_mapping_e, by = "input") %>%
  left_join(target_mapping_e, by = "sector") %>%
  group_by(scenario, Units, year, Source, Target) %>%
  summarize(value = sum(value)) %>%
  ungroup()
```

#### Electricity

```{r fig.width=8, warning=FALSE}
elec_energy_by_subsector <- elecEnergyInputBySubsector %>% 
  filter(Units == 'EJ') %>%
  filter(!input %in% c('backup_electricity', 'csp_backup'),
         !subsector %in% c("nuclear", "geothermal")) %>% #don't want to double count electricity from backup, and nuclear and geothermal are reported from output
  left_join(target_mapping_e, by = 'sector') %>% 
  left_join(source_mapping_e, by = 'input') 

#hydropower is only available as an output. In the "direct equivalent" reporting convention used here, input = output
hydro_power <- elecGenBySubsector %>%
  filter(subsector == 'hydro') %>%
  mutate(Source = 'Hydropower',
         Target = 'Electricity')

# nuclear's reported thermal inputs assume a 3:1 conversion, so for "direct equivalent" reporting we use the output
nuclear <- elecGenBySubsector %>%
  filter(subsector == 'nuclear') %>%
  mutate(Source = 'Nuclear',
         Target = 'Electricity') 

# geothermal's reported thermal inputs assume a 10:1 conversion, so for "direct equivalent" reporting we use the output
geothermal <- elecGenBySubsector %>%
  filter(subsector == 'geothermal') %>%
  mutate(Source = 'Geothermal',
         Target = 'Electricity')
```


```{r fig.width=8, warning=FALSE}
# put everything together
all_energy <- inputs_by_subsector_nonelec %>% 
  bind_rows(gas_processing_flows) %>% 
  bind_rows(backup) %>%
  bind_rows(elec_energy_by_subsector) %>% 
  bind_rows(hydro_power) %>% 
  bind_rows(nuclear) %>%
  bind_rows(geothermal)

Source_Target_all <- all_energy %>% 
  group_by(scenario, Units, Source, Target, year) %>%
  summarise(value = sum(value))  %>% 
  filter( Source != Target) %>% 
  filter( Target != 'Biomass') %>% 
  ungroup() 

datatable(Source_Target_all, filter = 'top', rownames = FALSE)
```


```{r fig.width=8, warning=FALSE}
# calculate losses 
# take the different of the sum of sources of a node and the sum of targets of a node and assign it to Losses target node. The Source would be the node it self. 

losses <- Source_Target_all %>%
  # cather data to have one column for "direction" (Source/Target) and one for "node"
  pivot_longer(cols = c(Source, Target), 
               names_to = "direction", 
               values_to = "node") %>%
  # only calculate for mid-tier/transformation flows
  left_join(node_mapping_e %>% filter(stage == "mid") %>% select("node" = "label", stage), by = "node" ) %>%
  filter(!is.na(stage)) %>%
  group_by(scenario, year, Units, node, direction) %>%
  summarize(total_value = sum(value), .groups = "drop") %>%
  pivot_wider(names_from = direction, values_from = total_value, values_fill = 0) %>%
  mutate(losses = Target - Source) %>%
  # filter(losses > 0) %>%
  # create the "Losses" rows with losses as target and the node as Source
  mutate(Source = node, Target = "Losses", value = losses) %>%
  select(scenario, Units, Source, Target, year, value) 

# add the losses back to the original dataset and complete it 
Source_Target_all_losses <- Source_Target_all %>% bind_rows(losses) %>% select(-Units) %>% 
  complete(scenario, year, nesting(Source, Target), fill = list(value = 0)) %>% mutate(units = "EJ")

```


```{r fig.width=8, warning=FALSE}
# energy losses plot 
if (F) {
  losses %>% 
  ggplot(aes(x = year, y = value, color = Source, linetype = Source)) +
  geom_line(size = 1) +
  scale_color_manual(values = node_mapping_e %>% filter(stage == "mid") %>% pull(hex)) +
  facet_wrap(~scenario, nrow = 2) +
  labs(title = "Energy Efficiency Losses: IM3 scenarios", x = "Year", y = "Losses (EJ)") +
  theme_bw() 
}

```

#### Plotting 

```{r fig.width=8, warning=FALSE}
scenario_name <- "rcp45cooler_ssp3"
plot_scenario_name <- 'RCP 4.5 Cooler SSP3'

select_year <- '2050'
gcam_data_unit <- 'EJ'

# sankey formatting
link_alpha <- .5

# source/target mapping

node_mapping_in <- node_mapping_e

# GCAM data
gcam_data <- Source_Target_all_losses %>% 
  filter(scenario == scenario_name) %>% filter( year == select_year) %>% select(-scenario)

all_links <- c(gcam_data$Source, gcam_data$Target) %>% unique()

node_mapping_e <- node_mapping_in %>% filter(label %in% all_links)

node_mapping_e$node <- 0:(nrow(node_mapping_e)-1)

# process node data
links_data <- gcam_data %>% 
  # filter(Source %in% c("Hydropower", "Solar")) %>% 
  select(Source, Target, value) %>% 
  # mutate(Target = ifelse(str_detect(Target, 'Ind'), 'Industry', Target)) %>% 
  group_by(Source, Target) %>% 
  summarize(value = sum(value)) %>% 
  ungroup() %>% 
  rename(Source_label = Source,
         Target_label = Target) %>% 
  left_join(node_mapping_e %>% select(label, node), by = c('Source_label' = 'label')) %>% 
  rename(Source_node = node) %>% 
  left_join(node_mapping_e %>% select(label, node), by = c('Target_label' = 'label')) %>% 
  rename(Target_node = node) %>% 
  left_join(node_mapping_e %>% select(label, stage, hex, color_name), by = c('Source_label' = 'label')) %>% 
  mutate(rgb = apply(FUN = paste, MARGIN = 2, X = col2rgb(hex), collapse = ',')) %>% 
  mutate(rgba = paste0('rgba(', rgb, ', ', link_alpha,')')) %>% 
  mutate(link_label = paste(Source_label, round(value, digits = 1),'EJ')) %>% 
  filter(value>0) %>% 
  arrange(Source_node)

datatable(links_data, filter = 'top', rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE))
```


```{r fig.width=8, warning=FALSE}
# process node percent labels

# source
source_sum <- links_data %>% 
  select(Source_label, value) %>% 
  left_join(node_mapping_e %>% select(label, stage), by = c('Source_label' = 'label')) %>% 
  rename(label=Source_label) %>% 
  filter(tolower(stage) == 'source') %>% 
  group_by(label, stage) %>% 
  summarize(node_sum = sum(value))

source_total <- source_sum %>% 
  pull(node_sum) %>% sum

source_percent <- source_sum %>% 
  mutate(percent = node_sum/source_total*100) %>% 
  left_join(node_mapping_e) %>% 
  arrange(node) %>% 
  mutate(x = .01) %>% 
  mutate(csum_norm = source_total)
source_percent$csum <- cumsum(source_percent$node_sum)
source_percent$start <- lag(source_percent$csum)

# target
target_sum <- links_data %>% 
  select(Target_label, value) %>% 
  left_join(node_mapping_e %>% select(label, stage), by = c('Target_label' = 'label')) %>% 
  rename(label=Target_label) %>% 
  filter(stage == 'target') %>% 
  group_by(label, stage) %>% 
  summarize(node_sum = sum(value))

target_total <- target_sum %>% 
  pull(node_sum) %>% sum

target_percent <- target_sum %>% 
  mutate(percent = node_sum/target_total*100) %>% 
  left_join(node_mapping_e) %>% 
  arrange(node) %>% 
  mutate(x = .95) %>% 
  mutate(csum_norm = target_total)
target_percent$csum <- cumsum(target_percent$node_sum)
target_percent$start <- lag(target_percent$csum)

# Intermediate Carriers Flows in
intermediate_nodes <- node_mapping_e %>% filter(stage == 'mid') %>% pull(label)
 intermediate_flows_in_total <- links_data %>%
   filter(Target_label %in% intermediate_nodes) %>% 
   group_by(Target_label) %>% 
   summarize(node_sum = sum(value))
 
intermediate_percent <- intermediate_flows_in_total %>% 
 rename(label = Target_label) %>% 
 mutate(stage = 'mid') %>% 
 mutate(percent =node_sum/source_total*100) %>% 
 left_join(node_mapping_e)

intermediate_total <- intermediate_percent %>% pull(node_sum) %>% sum

intermediate_flows_out_total <- links_data %>%
 filter(Source_label %in% intermediate_nodes) %>% 
 group_by(Source_label) %>% 
 summarize(value = sum(value))
  
```


```{r fig.width=8, warning=FALSE}
# process node locations 

# final node info
nodes_data <- bind_rows(source_percent, intermediate_percent, target_percent) %>%
  arrange(node) %>%
  replace_na(list(start = 0)) %>% 
  mutate(mid_point = (start+csum)/2) %>% 
  mutate(y = mid_point/csum_norm) %>% 
  mutate(y = ifelse(label == 'Gas', 0.5,
                    ifelse(label == 'Liquid Fuels', 0.2,
                    ifelse(label == 'Electricity', 0.6,
                    ifelse(label == 'Hydrogen',0.9,y))))) %>% 
  mutate(x = ifelse(label == 'Gas', 0.25,
                    ifelse(label == 'Liquid Fuels', 0.4,
                           ifelse(label == 'Electricity', 0.6,
                                  ifelse(label == 'Hydrogen', 0.7,x))))) %>%
  mutate(node_label = ifelse(is.na(node_sum), label, 
                               paste0(label, ' ',round(node_sum, digits = 1) , gcam_data_unit, 
                                      ' ', round(percent, digits = 1),'%'))) 
  

# Check that Source and Targets in Links are in the node mapping

if( any(is.na(links_data$Source_node)) ) stop("Check Source number mapping - NA's")
if( any(is.na(links_data$Target_node)) ) stop("Check Target number mapping - NA's")
  
datatable(nodes_data, filter = 'top', rownames = FALSE, options = list(pageLength = 20, scrollX = TRUE))
```


```{r fig.width=8, warning=FALSE}
# save files for Kendall
names(Source_Target_all_losses) <- tolower(names(Source_Target_all_losses)) 
write_csv(Source_Target_all_losses %>% select(scenario, source, target, year, value, units), paste0("../", data_dir, 'allenergy_source_target.csv'))
# write_csv(nodes_data, paste0("../", data_dir, 'allenergy_nodes_data.csv'))
# write_csv(links_data, paste0("../", data_dir, 'allenergy_links_data.csv'))

```


```{r fig.width=8, warning=FALSE}
# plot sankey
sankey_figure <- plot_ly( 
      type = "sankey",
      # arrangement = "snap",
      domain = list(x =  c(0,1),y =  c(0,1)),
      orientation = "h",
      valueformat = ".0f",
      valuesuffix = gcam_data_unit,

# Nodes  
      node = list( label = nodes_data %>% pull(node_label),
                   color = nodes_data %>% pull(hex),
                   x = nodes_data %>% pull(x),
                   y = nodes_data %>% pull(y),
                   pad = 3,
                   thickness = 15,
                   line = list(color = "black",width = 0.5)),
  
# Links
      link = list(source = links_data$Source_node,
                  target = links_data$Target_node,
                  value =  links_data$value,
                  color =  links_data$rgba)
) 

# add Formatting
plot_title <- paste0('Energy - ', plot_scenario_name, ' - ',select_year)
sankey_figure <- sankey_figure %>% layout(
  title = plot_title,
  font = list(size = 11),
  xaxis = list(showgrid = F, zeroline = F),
  yaxis = list(showgrid = F, zeroline = F))

sankey_figure


```



```{r fig.width=8, warning=FALSE}


```






