---
title: "Energy-Water Flows from the IM3 GCAM-USA Scenarios"
author: "Hassan Niazi (hassan.niazi@pnnl.gov)"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    toc: true
    # toc_float: TRUE
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE, warning=FALSE}
# by default collapse/hide the code
# knitr::opts_chunk$set(echo = FALSE)
# set working directory to one folder up
setwd("../")
# getwd()
source("./R/0_functions.R")
```

## For the EPRI's WET Forum

-   Goal: plot a water sankey, energy sankey, energy-water sankey (even if a partial one), show evolution over time
-   Let's hand-pick the relevant queries. Once finalized, we can use `rgcam` to query the data.

### Query and load IM3 data

Run this `query_im3_scen("epri")` only once to query from remote IM3 databases. Once a `.dat` file is created, we can load the existing project data by `loadProject(proj = "im3scen_epri.dat")`.

```{r message=FALSE, warning=FALSE}
# query the data
# im3_epri <- query_im3_scen("epri")
```

```{r, warning = FALSE}
# load the data
im3_epri <- loadProject(proj = paste0("../", data_dir, "im3scen_epri.dat"))
```

```{r}
# scenarios and queries 
listScenarios(im3_epri)
listQueries(im3_epri)
```

### Water Sankey

Overall required information

-   water availability by source (e.g., groundwater, surface water, desalination)
-   water use by categories (e.g., agriculture, industry, domestic, power etc)
-   water use by end-use sectors (e.g., crops, industry, electricity etc)

GCAM queries

-   availability
    -   resource supply curves
    -   basin level available runoff
    -   total runoff available
    -   total groundwater available
-   use by source
    -   resource production
    -   resource production by tech and vintage
    -   water withdrawals by water source (runoff vs. groundwater)
-   use by region/basin
    -   water withdrawals by region
-   use by category
    -   water withdrawals by water mapping source
    -   water withdrawals by tech
    -   water withdrawals by tech (all elec)
    -   water withdrawals by state, sector, basin (includes desal)

```{r, warning = FALSE}
# get queries 
# availability 
resourceSupplyCurves <- getQuery(im3_epri, "resource supply curves") %>% filter_CONUSregions()
basinLevelAvailableRunoff <- getQuery(im3_epri, "basin level available runoff") %>% filter_CONUSregions() 
# totalRunoffAvailable <- getQuery(im3_epri, "total runoff available") %>% filter_CONUSregions()
totalGroundwaterAvailable <- getQuery(im3_epri, "total groundwater available") %>% filter_CONUSregions()
# production 
resourceProduction <- getQuery(im3_epri, "resource production") %>% filter_CONUSregions()
resourceProductionByTechVintage <- getQuery(im3_epri, "resource production by tech and vintage") %>% filter_CONUSregions()
waterWithdrawalsByWaterSource <- getQuery(im3_epri, "water withdrawals by water source (runoff vs. groundwater)") %>% filter_CONUSregions()
# use by region/basin 
# waterWithdrawalsByRegion <- getQuery(im3_epri, "water withdrawals by region") %>% filter_CONUSregions()
waterWithdrawalsByStateSectorBasin <- getQuery(im3_epri, "water withdrawals by state, sector, basin (includes desal)") %>% filter_CONUSregions()
# use by category 
waterWithdrawalsByWaterMappingSource <- getQuery(im3_epri, "water withdrawals by water mapping source") %>% filter_CONUSregions()
waterWithdrawalsByTech <- getQuery(im3_epri, "water withdrawals by tech") %>% filter_CONUSregions()
# waterWithdrawalsByTechAllElec <- getQuery(im3_epri, "water withdrawals by tech (all elec)") %>% filter_CONUSregions()
```

```{r, messsage = T, warning = FALSE}
data_tables_water <- list(
  "resourceSupplyCurves" = resourceSupplyCurves,
  "basinLevelAvailableRunoff" = basinLevelAvailableRunoff,
  # "totalRunoffAvailable" = totalRunoffAvailable,
  "totalGroundwaterAvailable" = totalGroundwaterAvailable,
  "resourceProduction" = resourceProduction,
  "resourceProductionByTechVintage" = resourceProductionByTechVintage,
  "waterWithdrawalsByWaterSource" = waterWithdrawalsByWaterSource,
  # "waterWithdrawalsByRegion" = waterWithdrawalsByRegion,
  "waterWithdrawalsByStateSectorBasin" = waterWithdrawalsByStateSectorBasin,
  "waterWithdrawalsByWaterMappingSource" = waterWithdrawalsByWaterMappingSource,
  "waterWithdrawalsByTech" = waterWithdrawalsByTech
  # "waterWithdrawalsByTechAllElec" = waterWithdrawalsByTechAllElec
)

# print column names of each datatable
lapply(data_tables_water, function(x) colnames(x))

# print the first few rows of each datatable
lapply(data_tables_water, function(x) (x))
```

Let's process each piece to prepare the format of: scenario, source, target, year, value. Scenario and year could be filtered for each Sankey.

#### Water Availability

```{r fig.width=8, warning=FALSE}
unique(resourceSupplyCurves$resource)
unique(resourceSupplyCurves$subresource)

resourceSupplyCurves_water <- resourceSupplyCurves %>% filter(subresource %in% c("groundwater", "runoff")) %>% 
  filter(grade != "grade hist") %>% #exclude historical usage
  filter_basin_resource() %>%
  group_by(scenario, resource, subresource, year) %>% 
  summarise(value = sum(value)) %>% ungroup() %>% 
  remove_water_withdrawals_string() %>% 
  select(scenario, source = resource, target = subresource, year, value)

datatable(resourceSupplyCurves_water, filter = "top") 
plot_sankey(resourceSupplyCurves_water, "Water Availability by Basin and Resource Type")

```

Total resource availability doesn't change over time. This doesn't look right because basin level available runoff (which uses `max-annual-subresource` instead of summing all `grades`) is much more and changes over time.

<!-- -   *totalRunoffAvailable* and `totalGroundwaterAvailable` are the pre-filtered versions of `resourceSupplyCurves` (i.e., they filter for subresource just like we did above) -->
<!-- -   *totalRunoffAvailable* and `totalGroundwaterAvailable` are the same as `resourceSupplyCurves` but they are filtered for the USA region. -->
<!-- -   IMPORTANT: *totalRunoffAvailable* should NOT be used because runoff availability is specified by the `max-annual-subresource`, not the `grade` -->


```xml
<resourceQuery title="total groundwater available">
    <axis1 name="grade">grade</axis1>
    <axis2 name="Year">available</axis2>
    <xPath buildList="true" dataName="output" group="false" sumAll="true">*[@type='resource']/*[@type='subresource' and @name='groundwater']/grade/available/node()</xPath>
</resourceQuery>
```


`basinLevelAvailableRunoff` is the one that uses `max-annual-subresource` instead of summing all `grades`

```xml
<resourceQuery title="basin level available runoff">
    <axis1 name="Basin">resource[@name]</axis1>
    <axis2 name="Year">max-annual-subresource[@year]</axis2>
    <xPath buildList="true" dataName="input" group="false" sumAll="false">*[@type = 'resource' and contains(@name, 'water withdrawals')]/*[@type = 'subresource' and contains(@name, 'runoff')]/max-annual-subresource/node()</xPath>
</resourceQuery>
```

Let's use a combination of `basinLevelAvailableRunoff` and `totalGroundwaterAvailable` to get the total water availability by basin and year.

```{r fig.width=8, warning=FALSE}
totalWaterAvailability <- rbind(
  basinLevelAvailableRunoff %>% filter_basin_resource(basin) %>% remove_water_withdrawals_string(basin) %>% 
    # we may have to switch the source and target here to start from water sources and go to basins
    select(scenario, target = basin, source = subresource, year, value),
  totalGroundwaterAvailable %>% 
    filter_basin_resource() %>% remove_water_withdrawals_string() %>% 
    filter(grade != "grade hist") %>% 
    group_by(scenario, resource, subresource, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    select(scenario, target = resource, source = subresource, year, value)
)

datatable(totalWaterAvailability, filter = "top")
plot_sankey(totalWaterAvailability, "Total Water Availability by Basin and Resource Type")

```

This looks better! Let's move to water withdrawals.

#### Water Withdrawals

Let's start by quantifying water withdrawals by water source (runoff vs. groundwater) for each basin.

- `resourceProduction` gives ALL resources, would need to filter by `_water withdrawals`. BUT this will give TOTAL basin level production (as basins are *resources*) and not by water source (runoff vs. groundwater) which are *subresources*.
- `resourceProductionByTechVintage` is the most detailed resource query, with information for regions, resources (water withdrawals), subresources (runoff, groundwater), production by year. 

But let's use a ready-made query `waterWithdrawalsByWaterSource` which gives water withdrawals by water source (runoff vs. groundwater) for each basin (resource).

- However, this still doesn't have desalination. `resourceProduction` has desalination, but on the USA level. We will need to get desalination from `waterWithdrawalsByStateSectorBasin`. 


```{r fig.width=8, warning=FALSE}
# prepare desalination tables 
desalByBasin <- waterWithdrawalsByStateSectorBasin %>% 
    filter(region != "USA") %>% 
    filter(technology == "desalination") %>%
    filter_CONUSregions(subsector, basins_conus_resource) %>% 
    group_by(scenario, subsector, technology, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    select(scenario, source = technology, target = subsector, year, value)

desalByUseCategory <- waterWithdrawalsByStateSectorBasin %>% 
    filter(region != "USA") %>% 
    filter(technology == "desalination") %>% 
    filter_CONUSregions(subsector, basins_conus_resource) %>% 
    group_by(scenario, sector, technology, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    select(scenario, source = technology, target = sector, year, value)

desalByState <- waterWithdrawalsByStateSectorBasin %>%
    filter(region != "USA") %>% 
    filter(technology == "desalination") %>% 
    filter_CONUSregions(subsector, basins_conus_resource) %>%   
    group_by(scenario, region, technology, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    select(scenario, source = technology, target = region, year, value)


plot_sankey(rbind(totalWaterAvailability, desalByBasin), "Total Water Availability by Basin and Resource Type")
```

```{r fig.width=8, warning=FALSE}
df_waterWithdrawalsByWaterSource <- rbind(
  waterWithdrawalsByWaterSource %>% 
    filter_basin_resource() %>% 
    remove_water_withdrawals_string() %>%
    select(scenario, source = subresource, target = resource, year, value)
  , 
  desalByBasin
  )

datatable(df_waterWithdrawalsByWaterSource, filter = "top")
plot_sankey(df_waterWithdrawalsByWaterSource, "Water Withdrawals by Basin and Water Source")

```
- Only 7 basins use groundwater in 2050 example above. 

#### Water Withdrawals by State, Sector, and Basin
```{r fig.width=8, warning=FALSE}

df_waterWithdrawalsByStateSectorBasin <- waterWithdrawalsByStateSectorBasin %>% 
  filter(region != "USA") %>% # exclude USA level data because it aggregated up from basin level data, but lacks the technology detail (name of the basin here)
  filter_CONUSregions(technology, basins_conus_resource) %>% # not sure why nonCONUS basins have crept in here at the technology level, but remove them 
  filter(technology != "desalination") %>% # we have desalination at the basin level in the previous plot 
  replace_after_irr_string() %>% # if this is removed, each basin will supply to it's own irrigation demand
  group_by(scenario, sector, technology, year) %>%
  summarise(value = sum(value)) %>% ungroup() %>%
  select(scenario, source = technology, target = sector, year, value)

datatable(df_waterWithdrawalsByStateSectorBasin, filter = "top")
plot_sankey(df_waterWithdrawalsByStateSectorBasin, "Water Withdrawals by Use Category and Basin")

# let's try to piece this with waterWithdrawalsByWaterSource to the the water source type, basin, water use type flow
plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource), "Water Withdrawals by Use Category and Basin")

```
TODO: Q: What is going on with Great Lakes? Why is it showing up in the water withdrawals but not in the water availability? 
```{r fig.width=8, warning=FALSE}
# exclude great lakes for now 
# we brought it back by borrowing from Canada. 
# df_waterWithdrawalsByStateSectorBasin <- df_waterWithdrawalsByStateSectorBasin %>% filter(source != "Great Lakes")

# plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource), "Water Withdrawals by Use Category and Basin")
```


#### Detailed Water Use 
Let's develop the water use categories further.

- both `waterWithdrawalsByTech` and `waterWithdrawalsByTechAllElec` are the same queries. 
- get short basin names `unique((waterWithdrawalsByTech_watCat %>% filter(grepl("biomassTree", subsector)))$subsector)`

```{r fig.width=8, warning=FALSE}
waterWithdrawalsByTech_watCat <- waterWithdrawalsByTech %>% map_water_use_to_categories() %>% filter_CONUSregions(region, states_conus)

watcategory_sector_use <- waterWithdrawalsByTech_watCat %>% 
  group_by(scenario, sector, watcategory, year) %>%
  summarise(value = sum(value)) %>% ungroup() %>%
  select(scenario, source = watcategory, target = sector, year, value)


plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource, watcategory_sector_use), "Water Withdrawals by Use Category and Basin")

# use by subsector: same plot as previous after aggregation 
# watcategory_subsector_use <- waterWithdrawalsByTech_watCat %>% 
#   remove_GLUnames(subsector) %>% 
#   group_by(scenario, subsector, watcategory, year) %>%
#   summarise(value = sum(value)) %>% ungroup() %>%
#   select(scenario, source = watcategory, target = subsector, year, value)

# this is essentially the same as the previous plot
# plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource, watcategory_subsector_use), "Water Withdrawals by Use Category and Basin")
```

```{r fig.width=8, warning=FALSE}
# expand electricity water use 
watcategory_subsector_use <- waterWithdrawalsByTech_watCat %>% 
  filter(sector == "electricity") %>%
  # remove_GLUnames(subsector) %>%
  group_by(scenario, sector, subsector, year) %>%
  summarise(value = sum(value)) %>% ungroup() %>%
  select(scenario, source = sector, target = subsector, year, value)

plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource, watcategory_sector_use, watcategory_subsector_use), "Water Withdrawals by Use Category and Basin")
```


Further technology level disaggregation for electricity water use 
```{r fig.width=9, warning=FALSE}
waterEnduseByTech <- waterWithdrawalsByTech_watCat %>% 
  filter(sector == "electricity") %>%
  clean_cooling_tech("technology", cooling_techs, cooling_techs_years) %>% 
  group_by(scenario, subsector, technology, year) %>%
  summarise(value = sum(value)) %>% ungroup() %>%
  select(scenario, source = subsector, target = technology, year, value)

plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource, watcategory_sector_use, watcategory_subsector_use, waterEnduseByTech), "Water Use by Source, Basin, Category, Sector, and Technology, and Basin")

```

#### Water Requirement for Hydro

``` {r fig.width=9, warning=FALSE}
# add water requirement for hydro in the water diagram 
# ideally use hydro generation and create a water flow dynamically for all scenarios all years using the 3060 km3/EJ (853 MG/BBTU) coefficient 
# but right now, I am just going to create a single value

# hydro water requirement 

plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource, watcategory_sector_use, 
                  watcategory_subsector_use %>% 
                    # add hydro for rcp45cooler_ssp3 scenario and year 2050
                    add_row(scenario = "rcp45cooler_ssp3", source = "electricity", target = "hydro", year = 2050, value = 0.995 * 3060),
                  waterEnduseByTech) %>% rename_water_use_categories() %>% 
              mutate(source = ifelse(source == "runoff", "surface water", source)) %>% 
              mutate(target = ifelse(target == "industry", "industry (elec excluded)", target)),
            "Water Use by Source, Basin, Category, Sector, and Technology, and Basin")
  
plot_sankey(rbind(df_waterWithdrawalsByStateSectorBasin, df_waterWithdrawalsByWaterSource, watcategory_sector_use, 
                  watcategory_subsector_use %>% 
                    # add hydro for rcp45cooler_ssp3 scenario and year 2050
                    add_row(scenario = "rcp45cooler_ssp3", source = "electricity", target = "hydro", year = 2050, value = 0.995 * 3060 * 1e-3),
                  waterEnduseByTech) %>% rename_water_use_categories() %>% 
              mutate(source = ifelse(source == "runoff", "surface water", source)) %>% 
              mutate(target = ifelse(target == "industry", "industry (elec excluded)", target)),
            "Water Use by Source, Basin, Category, Sector, and Technology, and Basin")
  

# TODO: 
# labels
# time
# scenarios
# industry (elec excluded)
# relabel td sectors 
# runoff = surface water 
# annotations for levels (basins: HUC2)
# can colors the map as the plotly basins 
# avoid repeating colors
```


``` {r fig.width=9, warning=FALSE}
elecGenByGenTechCoolingTech_gUSA <- getQuery(im3_epri, "elec gen by gen tech and cooling tech (incl cogen)") %>% filter_CONUSregions()
elecGenByGenTechCoolingTech <- getQuery(im3_epri, "elec gen by gen tech and cooling tech") %>% filter_CONUSregions()
hydro_water_req <- elecGenByGenTechCoolingTech %>%
    filter(subsector == "hydro") %>%
    group_by(scenario, subsector, output, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>% 
  mutate(source = "water_td_elec_W",
         hydro_water_req = value * 3060 * 1e-3) %>%
  select(scenario, source, target = subsector, year, value = hydro_water_req)
```

## Water Diagram for USA

```{r fig.width=10, warning=FALSE}
# remove basins and electricity aggregations (only keep technologies)

# 1. prepare resources 

desal_totalUSA <- waterWithdrawalsByStateSectorBasin %>% 
    filter(region != "USA") %>% 
    filter(technology == "desalination") %>% 
    group_by(scenario, sector, technology, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    select(scenario, source = technology, target = sector, year, value)

# create surface to groundwater shares by basin and apply those to use categories to determine amount of surface and groundwater used for each demand 
# 
# this calculates availabilty shares (which are not really relevant becasue we need withdrawal shares)
# totalWaterAvailability_USA <- rbind(
#   basinLevelAvailableRunoff %>% filter_basin_resource(basin) %>% remove_water_withdrawals_string(basin) %>% 
#     # we may have to switch the source and target here to start from water sources and go to basins
#     select(scenario, target = basin, source = subresource, year, value),
#   totalGroundwaterAvailable %>% 
#     filter_basin_resource() %>% remove_water_withdrawals_string() %>% 
#     filter(grade != "grade hist") %>% 
#     group_by(scenario, resource, subresource, year) %>%
#     summarise(value = sum(value)) %>% ungroup() %>%
#     select(scenario, target = resource, source = subresource, year, value)
# ) %>% group_by(scenario, source, year) %>%
#     summarise(value = sum(value)) %>% ungroup() 


waterWithdrawalsByWaterSource_shares_bybasin <- waterWithdrawalsByWaterSource %>% 
  filter_basin_resource() %>%
  remove_water_withdrawals_string() %>%
  group_by(scenario, region, resource, year) %>% mutate(share = value / sum(value)) %>% ungroup()

datatable(waterWithdrawalsByWaterSource_shares_bybasin, filter = "top")


waterWithdrawalsByStateSectorBasin_source <- waterWithdrawalsByStateSectorBasin %>% 
  # filter(scenario == "rcp85cooler_ssp5", year == "2050", sector == "water_td_an_W") %>%
  filter(region != "USA") %>% # exclude USA level data because it aggregated up from basin level data, but lacks the technology detail (name of the basin here)
  filter_CONUSregions(technology, basins_conus_resource) %>% # not sure why nonCONUS basins have crept in here at the technology level, but remove them 
  filter(technology != "desalination") %>% # we have desalination at the basin level in the previous plot 
  replace_after_irr_string() %>% # if this is removed, each basin will supply to it's own irrigation demand
  group_by(scenario, sector, technology, year) %>%
  summarise(value = sum(value)) %>% ungroup() %>%
  select(scenario, source = technology, target = sector, year, demand_withdraw = value) %>% 
  left_join(waterWithdrawalsByWaterSource_shares_bybasin, by = c("scenario", "source" = "resource", "year")) %>% 
  mutate(source_disagg = demand_withdraw * share) 

datatable(waterWithdrawalsByStateSectorBasin_source, filter = "top")
```


```{r fig.width=10, warning=FALSE}
# create 3 way shares among runoff gw and desal by using  desal from waterWithdrawalsByStateSectorBasin and runoff and gw from waterWithdrawalsByWaterSource
gwRunoffDesalShares <- waterWithdrawalsByWaterSource %>% 
  filter_basin_resource() %>%
  remove_water_withdrawals_string() %>% select(!c("Units", "region")) %>% 
  rbind(waterWithdrawalsByStateSectorBasin %>% 
          filter(region != "USA") %>% 
          filter(technology == "desalination") %>% 
          group_by(scenario, subsector, technology, year) %>%
          summarise(value = sum(value)) %>% ungroup() %>%
          select(scenario, subresource = technology, resource = subsector, year, value)
  ) %>% group_by(scenario, resource, year) %>% mutate(share = value / sum(value)) %>% ungroup()

waterWithdrawalsByStateSectorBasin_source_3 <- waterWithdrawalsByStateSectorBasin %>% 
  # filter(scenario == "rcp85cooler_ssp5", year == "2050", sector == "water_td_an_W") %>%
  filter(region != "USA") %>% # exclude USA level data because it aggregated up from basin level data, but lacks the technology detail (name of the basin here)
  filter_CONUSregions(technology, basins_conus_resource) %>% # not sure why nonCONUS basins have crept in here at the technology level, but remove them 
  filter(technology != "desalination") %>% # we have desalination at the basin level in the previous plot 
  replace_after_irr_string() %>% # if this is removed, each basin will supply to it's own irrigation demand
  group_by(scenario, sector, technology, year) %>%
  summarise(value = sum(value)) %>% ungroup() %>%
  select(scenario, source = technology, target = sector, year, demand_withdraw = value) %>% 
  left_join(gwRunoffDesalShares, by = c("scenario", "source" = "resource", "year")) %>% 
  mutate(source_disagg = demand_withdraw * share) 


waterSourceUseCategories3 <- waterWithdrawalsByStateSectorBasin_source_3 %>% 
  group_by(scenario, subresource, target, year) %>% 
  summarise(value = sum(source_disagg)) %>% ungroup() %>% 
  select(scenario, source = subresource, target, year, value) %>% 
  # note we are aggregating up basin-level irrigation from desalination here 
  rbind(desal_totalUSA %>% replace_after_irr_string(target) %>% 
          group_by(scenario, source, target, year) %>% summarise(value = sum(value)) %>% ungroup()
        ) 

plot_sankey(waterSourceUseCategories3, "Water Source and Use Category")  
plot_sankey(rbind(waterSourceUseCategories3, waterCateogryTech), "Water Source, Use Category, and End-Use")

# CONLCUSION: THIS IS EXACTLY THE SAME AS THE ONE PREPARE WIHTOUT 3 WAY SHARES                              

```

```{r fig.width=10, warning=FALSE}
# US level supply and demand
waterSourceUseCategories <- waterWithdrawalsByStateSectorBasin_source %>% 
  group_by(scenario, subresource, target, year) %>% 
  summarise(value = sum(source_disagg)) %>% ungroup() %>% 
  select(scenario, source = subresource, target, year, value) %>% 
  # note we are aggregating up basin-level irrigation from desalination here 
  rbind(desal_totalUSA %>% replace_after_irr_string(target) %>% 
          group_by(scenario, source, target, year) %>% summarise(value = sum(value)) %>% ungroup()
        ) 

plot_sankey(waterSourceUseCategories, "Water Source and Use Category")   

```

```{r fig.width=10, warning=FALSE}
# water source, use category, and end-use 
waterCateogryTech <- 
  # we will use different details for each use category 
  rbind(
    # electricity on tech level
    waterWithdrawalsByTech_watCat %>% 
      filter(watcategory == "water_td_elec_W") %>% 
      clean_cooling_tech("technology", cooling_techs, cooling_techs_years) %>%
      group_by(scenario, watcategory, technology, year) %>%
      summarise(value = sum(value)) %>% ungroup() %>% 
      select(scenario, source = watcategory, target = technology, year, value)
    , 
    # everything else on sector level 
    waterWithdrawalsByTech_watCat %>% 
      filter(watcategory != "water_td_elec_W") %>% 
      group_by(scenario, watcategory, sector, year) %>%
      summarise(value = sum(value)) %>% ungroup() %>%
      select(scenario, source = watcategory, target = sector, year, value)
  )

plot_sankey(rbind(waterSourceUseCategories, waterCateogryTech), "Water Source, Use Category, and End-Use")
```


```{r fig.width=10, warning=FALSE}
# water source, basin, use category, and end-use
plot_sankey(rbind(df_waterWithdrawalsByWaterSource, df_waterWithdrawalsByStateSectorBasin, waterCateogryTech, hydro_water_req), "Water Source, Basin, Use Category, and End-Use")
```


```{r fig.width=10, warning=FALSE}
# SWAPPED basins: water source, basin, use category, and end-use
water_source_target <- rbind(df_waterWithdrawalsByWaterSource %>% swap_source_target(), 
                  waterSourceUseCategories,
                  waterCateogryTech, hydro_water_req) %>% 
  complete(scenario, year, nesting(source, target), fill = list(value = 0)) %>% mutate(units = "km3") %>% 
  select(scenario, source, target, year, value, units)

plot_sankey(water_source_target, "Water Source, Basin, Use Category, and End-Use")

write_csv(water_source_target , paste0("../", data_dir, "water_source_target.csv"))
```




```{r fig.width=10, warning=FALSE}
# water source, basin, use category, and end-use WITH BETTER COLORS 

waterSourceBasinsCategoriesUse_colors <- read_csv(paste0("../", data_dir, "waterSourceBasinsCategoriesUse.csv"))

plot_sankey(waterSourceBasinsCategoriesUse_colors, "Water Source, Basin, Use Category, and End-Use")
```


```{r fig.width=10, warning=FALSE}
# Combine source and target to get unique nodes and their colors
nodes <- waterSourceBasinsCategoriesUse_colors %>%
  select(source, source_color) %>%
  rename(node = source, node_color = source_color) %>%
  bind_rows(
    waterSourceBasinsCategoriesUse_colors %>%
      select(target, target_color) %>%
      rename(node = target, node_color = target_color)
  ) %>%
  distinct()

# Add color directly to node labels
node_labels <- unique(c(waterSourceBasinsCategoriesUse_colors$source, 
                        waterSourceBasinsCategoriesUse_colors$target))

# Map node colors to node labels
node_colors <- nodes$node_color[match(node_labels, nodes$node)]
```


```{r fig.width=10, warning=FALSE}
unique(waterSourceBasinsCategoriesUse_colors[c("source", "source_color")])$source_color[match(node_labels, unique(waterSourceBasinsCategoriesUse_colors$source))]


```












## Energy Sankey

Overall information required:

- resource production (coal, gas, oil etc)
- energy inputs to electricity generation 
    - input is resource 
    - 
- electricity generation 
    - 
- electricity use by categories 
    - 
- electricity use by end uses
    - 


Queries: 

- `elec energy input by elec gen tech and cooling tech`
- `elec water withdrawals by gen tech and cooling tech`
- `elec gen by gen tech and cooling tech (incl cogen)`
- `elec energy input by elec gen tech`
- `elec gen by gen tech and cooling tech`
- `elec water withdrawals by gen tech`
- `elec td inputs and outputs`
- `elec consumption by demand sector`


```{r fig.width=8, warning=FALSE}
# read queries 
# GCAM-USA specific queries 
elecEnergyInputByElecGenTechCoolingTech_gUSA <- getQuery(im3_epri, "elec energy input by elec gen tech and cooling tech") %>% filter_CONUSregions()
elecWaterWithdrawByElecGenTechCoolingTech_gUSA <- getQuery(im3_epri, "elec water withdrawals by gen tech and cooling tech") %>% filter_CONUSregions()
elecGenByGenTechCoolingTech_gUSA <- getQuery(im3_epri, "elec gen by gen tech and cooling tech (incl cogen)") %>% filter_CONUSregions()
# generic GCAM queries
elecEnergyInputByElecGenTech <- getQuery(im3_epri, "elec energy input by elec gen tech") %>% filter_CONUSregions()
elecGenByGenTechCoolingTech <- getQuery(im3_epri, "elec gen by gen tech and cooling tech") %>% filter_CONUSregions()
elecWaterWithdrawByGenTech <- getQuery(im3_epri, "elec water withdrawals by gen tech") %>% filter_CONUSregions()
elecTDInputsOutputs <- getQuery(im3_epri, "elec td inputs and outputs") %>% filter_CONUSregions()
elecConsumptionByDemandSector <- getQuery(im3_epri, "elec consumption by demand sector") %>% filter_CONUSregions()
```

```{r, messsage = T, warning = FALSE}
data_tables_ew <- list(
  "elecEnergyInputByElecGenTechCoolingTech_gUSA" = elecEnergyInputByElecGenTechCoolingTech_gUSA,
  "elecWaterWithdrawByElecGenTechCoolingTech_gUSA" = elecWaterWithdrawByElecGenTechCoolingTech_gUSA,
  "elecGenByGenTechCoolingTech_gUSA" = elecGenByGenTechCoolingTech_gUSA,
  "elecEnergyInputByElecGenTech" = elecEnergyInputByElecGenTech,
  "elecGenByGenTechCoolingTech" = elecGenByGenTechCoolingTech,
  "elecWaterWithdrawByGenTech" = elecWaterWithdrawByGenTech,
  "elecTDInputsOutputs" = elecTDInputsOutputs,
  "elecConsumptionByDemandSector" = elecConsumptionByDemandSector
)

# print column names of each datatable
lapply(data_tables_ew, function(x) colnames(x))

# print the first few rows of each datatable
lapply(data_tables_ew, function(x) (x))
```


```{r fig.width=10, warning=FALSE}
elec_source_target <- rbind(
  # water withdrawals by electricity generation technology
  elecWaterWithdrawByGenTech %>% 
    # mutate(value = value * 1e-1) %>% # change to x10 km3 for better visuals
    clean_cooling_tech("technology", cooling_techs, cooling_techs_years) %>%
    group_by(scenario, technology, input, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>% 
    select(scenario, source = input , target = technology, year, value)
  ,
  elecEnergyInputByElecGenTech %>% 
    clean_cooling_tech("technology", cooling_techs, cooling_techs_years) %>%
    group_by(scenario, technology, input, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>% 
    select(scenario, source = input , target = technology, year, value)
  ,
  # electricity generation by technology
  elecGenByGenTechCoolingTech %>% 
    clean_cooling_tech("technology", cooling_techs, cooling_techs_years) %>%
    group_by(scenario, subsector, technology, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>% 
    select(scenario, source = technology, target = subsector, year, value)
  ,
  # bridge gen sector to electricity supply sector
  elecGenByGenTechCoolingTech %>%
    # sum by subsector
    group_by(scenario, subsector, output, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    rename(source = subsector, target = output) 
  ,
  # inputs to the end use sectors
  elecTDInputsOutputs %>%
    filter(grepl("elect_", sector)) %>%
    group_by(scenario, sector, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    # TODO: add electricity_net_ownuse, and create a losses "end-use"
    mutate(source = "electricity") %>%
    select(scenario, source, target = sector, year, value)
  ,
  # electricity consumption by demand sector
  elecConsumptionByDemandSector %>%
    group_by(scenario, sector, input, year) %>%
    summarise(value = sum(value)) %>% ungroup() %>%
    select(scenario, source = input, target = sector, year, value)
) %>% filter(source != target)
```


```{r fig.width=10, warning=FALSE}
plot_sankey(elec_source_target, "Electricity Generation and Consumption")
```


```{r fig.width=10, warning=FALSE}
# remove water from electricity diagram 
plot_sankey(elec_source_target %>% filter(source != "water_td_elec_W"), yr = 2050, "Electricity Generation and Consumption")
```






























